---
title: "Recitation 10: SLR Inference"
subtitle: "SOLUTIONS"
author: "Haley Grant"
output:
  html_document:
    df_print: paged
    toc: yes
    number_sections: no
    toc_depth: '4'
    code_folding: show
---

```{r setup, include=FALSE}
# this is a set-up chunk
# DON'T CHANGE THESE SETTINGS--I'VE SET THEM UP TO MAKE THINGS RUN SMOOTHLY 
knitr::opts_chunk$set(echo = TRUE, error = TRUE, 
                      root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Set-up

## Load packages

```{r}
library(tidyverse)
if(!require(ggExtra)){install.packages("ggExtra")
  library(ggExtra)}

```

## Import data

```{r}

# import the data using the .RData file provided
load("Data/liver_data.RData", verbose = T)

# update units for LV
liver_data = liver_data %>% 
  rename(liver_volume_mL = liver_volume) %>%
  mutate(liver_volume = liver_volume_mL/1000)

```

# Simple Linear Regression

## Visualization and model from last time

Here I have reproduced the scatterplot and model we fit last time relating liver volume to age within the sample:

```{r}
# visualize
liver_data %>%
  ggplot(aes(x = age, y = liver_volume)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Age",y = "Liver Volume")

# fit model
model <- lm(liver_volume ~ age, data = liver_data)
# summary of model
summary(model)

# add standard error to this 
liver_data %>%
  ggplot(aes(x = age, y = liver_volume)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Age",y = "Liver Volume") + 
  geom_smooth(method = "lm", se = F)

```

## Check conditions

To trust our inference, we need to check the regression assumptions.

**Check the conditions for regression using residual plots. Do the regression assumptions appear to be (reasonably) met?**

L = linearity I = independence N = normality of residuals (or outcome is normally distributed at each value of X) E = equal variance

```{r}
# residual plots
# residual vs fitted values
# we can use this to check equal variance and linearity
plot(model, which = 1)

# qq plot of residuals
plot(model, which = 2)

```

Interpretation: Based on the residuals vs fitted values plots, linearity looks reasonable and equal variance looks alright. Based on the normal QQ plot, we see there are a few upper outliers (large positive residuals), but they look generally normal.

Here are some examples where the assumptions are clearly violated:

```{r}
# simulating data where assumptions are violated
set.seed(111)
x = runif(n = 200)
y1 = 2*x - 2*x^2 + rnorm(n = 200, sd = 0.1)
y2 = 2*x + rnorm(n = 200, sd = x)
y3 = 2*x + rexp(n = 200)

# residuals vs fitted values (nonlinearity)
plot(lm(y1 ~ x), which = 1)

# residuals vs fitted values (non equal variance)
plot(lm(y2 ~ x), which = 1)

# qq plot of residuals (non-normality)
plot(lm(y3 ~ x), which = 2)
```

# Simulation

Next we're going to run a simulation to help illustrate the difference between normality of the outcome variable (Y) in linear regression and normality of the residuals.

## Simulate data

Here I'm simulating some data and fitting a linear model on the fitted data. Run this chunk.

```{r}
set.seed(123)
model = lm(liver_volume ~ age, data = liver_data)
# simulate x
x = c(runif(n = 100, min = 5, max = 10), liver_data$age + 10)
# simulate y using model we fit earlier + add random (normal) noise
y = predict(model, newdata = data.frame(age = x)) + rnorm(n = length(x),mean = 0, sd = sqrt(sum(model$residuals^2)/model$df.residual)/2)

# fit model
model.sim = lm(y ~ x)
sim.data = data.frame(x, y , residuals = model.sim$residuals) 


```

## Plot y

**Make a histogram and QQ-plot of the simulated outcome variable, y. Does this variable look normally distributed?**

```{r}
# histogram of y
sim.data %>%
  ggplot(aes(x = y)) + 
  geom_histogram(color = "black", fill = "lightblue") + 
  theme_bw()

# qqplot of y
sim.data %>%
  ggplot(aes(sample = y)) + 
  geom_qq() + 
  geom_qq_line() + 
  labs(x = "Theoretical Normal Quantiles",
       y = "Observed Quantiles of Outcome Variable")

```

## Visualize Residuals

**Make a histogram and QQ-plot of the residuals from the linear model we fit on the simualted data. Does this variable look normally distributed?**

```{r}
# histogram of residuals
sim.data %>%
  ggplot(aes(x = residuals)) + 
  geom_histogram(color = "black", fill = "firebrick1") + 
  theme_bw()

# qqplot of residuals
sim.data %>%
  ggplot(aes(sample = residuals)) + 
  geom_qq() + 
  geom_qq_line() + 
  labs(x = "Theoretical Normal Quantiles",
       y = "Observed Quantiles of Residuals")
```

## Plots showing joint and marginal distributions

These plots are meant to help explain what is happening above. Notice that even though the residuals are normally distributed, the outcome variable is not because the x variable isn't symmetric.

Note: you won't ever need to make plots like these on your assignments, I'm just including these plots with the marginal distributions plotted to help explain this phenomenon.

### Plot y by x

```{r}

# scatterplot of y ~ x
ggMarginal(sim.data %>%
  ggplot(aes(x = x, y= y)) + 
  geom_point() + 
  theme_bw() + 
    geom_smooth(method = "lm", se = F), type = "histogram")

```

### Plot residuals by x

```{r}
# scatterplot of residuasl  ~ x
ggMarginal(sim.data %>%
  ggplot(aes(x = x, y= residuals)) + 
  geom_point(color = "red") + 
  theme_bw() + 
    geom_smooth(method = "lm", se = F), type = "histogram")

```

## Coefficient of determination

**What is the coefficient of determination (R squared) for the model we fit on the liver data? Interpret this value.**

```{r}
summary(model)$r.squared
```

Interpretation: 23.7% of the variation in liver volume can be "explained" by age (by the linear regression line using age as a predictor).

## Decomposing Variability

```{r}
n = 372
# variability in outcome
var(liver_data$liver_volume)
var(liver_data) * (n - 1)
ybar = mean(liver_data$liver_volume)
SST = sum((liver_data$liver_volume - ybar)^2)
SST

# variability in predicted values
yhatbar = mean(model$fitted.values) # yhatbar = ybar
SSR = sum((model$fitted.values - yhatbar)^2)
SSR

# variability in residuals
ebar = mean(model$residuals) # ebar ~= 0
SSE = sum((model$residuals - ebar)^2)
SSE

# calculate R2 by hand
# proportion variability explained
SSR / SST

# 1 - proportion variability not explained
 1- SSE / SST

# model-based R2
summary(model)$r.squared
```

## Hypothesis test for slope

### Hypotheses

**What are the null and alternative hypotheses for the test for the slope shown in the model summary output?**

Null: Slope in the population is 0 (i.e. liver volume and age have no linear association).

Alternative: Slope in the population is not 0 (i.e. liver volume and age have a linear association).

### Results

**What are the test statistic and p-value for this test. Interpret the results of this hypothesis test (use alpha = 0.05).**

Test statistic: T = (beta1hat - 0) / SE(beta1hat) = -10.73

p-value: 1.4 e-23 (p \< 2e-16)

Interpretation: reject null. we have sufficient evidence of a negative linear relationship between age and liver volume

```{r}
# exact calculations (by hand):
# sample size
n = 372
# slope estimate
betahat = model$coefficients[2]
# standard error
se = summary(model)$coefficients[2,"Std. Error"]
# test stat
t_obs = (betahat - 0) / se
t_obs
# p-value
pv = pt(abs(t_obs), df = n - 2, lower.tail = F) * 2
pv
```

## Inference for parameters

**Write some code to get the estimated 95% confidence intervals for each parameter and interpret the confidence intervals below.**

```{r}
# confidence interval for parameters
confint(model)

```

Interpretations:

Intercept: We are 95% confident the average liver volume for a 0-year-old is between 1.78 and 2.09 liters (this is probably wrong--extrapolation!!)

Slope: We are 95% confident the average difference in liver volume associated with a one-year increase in age is between -0.0145 and -0.01.

## Inference for average liver volume at specific ages

**Write some code to obtain 95% confidence intervals for the average liver volume at age 30, 50, 65, 80, and 90.**

```{r}

# confidence interval for estimated average liver volume at different ages
# for age = c(30, 50, 65, 80, 90)
predict(model, newdata = data.frame(age = c(30, 50, 65, 80, 90)),
        interval = "confidence")


```

Here we can visualize the confidence interval for the estimated liver volume across the range of ages observed in our sample:

```{r}
# visualization
liver_data %>%
  ggplot(aes(x = age, y = liver_volume)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Age",y = "Liver Volume") + 
  geom_smooth(method = "lm")


```

## Interpretation of confidence intervals

**What is the interpretation of these confidence intervals? What value should intervals like these contain 95% of the time?**

Answer: We are 95% confident that the **AVERAGE** liver volume among people who are 30, 50, 65, 80, and 90 is between lower and upper values given above, respectively.

## Inference for individual values of liver volume at specific ages

**Write some code to obtain 95% prediction intervals for an individual's liver volume of age 30, 50, 65, 80, and 90.**

```{r}
# prediction intervals for estimated  liver volume at different ages
# for age = c(30, 50, 65, 80, 90)
predict(model, newdata = data.frame(age = c(30, 50, 65, 80, 90)),
        interval = "prediction")


```

## Interpretation of prediction intervals

**What is the interpretation of these prediction intervals? What value should intervals like these contain 95% of the time?**

Answer: We are 95% confident that **an individual** who is 30, 50, 65, 80, and 90 will have a liver volume between lower and upper values given above, respectively.
